Explainability and Adversarial Robustness for RNNs

### 2019-12-19 paper/paper.tex
Small text changes

\glspl{rnn} yield attractive properties for %performing intrusion detection on network data.
constructing \glspl{ids} for network data.
With the rise of ubiquitous \gls{ml} systems, malicious actors have been catching up quickly to find new ways to exploit \gls{ml} for profit. Recently developed adversarial \gls{ml} techniques focus on computer vision and their applicability to network traffic is not straightforward: Network packets expose fewer features than an image, are sequential and impose several constraints on their features. %For example, destination ports cannot be modified since that could impair communication.

We show that despite these completely different characteristics, adversarial samples can be reliably generated for \glspl{rnn}.
To understand a classifier's potential for misclassification, we extend existing explainability techniques and propose new ones, suitable particularly % specifically crafted %AH most can also be used for non-sequential models, so lets not restrict applicability too much
for sequential data. Applying them shows that already the first packets of a communication flow are of crucial importance and are likely to be targeted by attackers. Feature importance methods show that even relatively unimportant features can be effectively abused to generate adversarial samples.
Since traditional evaluation metrics such as accuracy are not sufficient for quantifying the adversarial threat, we propose the \gls{ars} for comparing \glspl{ids}, capturing a common notion of adversarial robustness, and show that an adversarial training procedure can significantly and successfully reduce the attack surface.% and improve \gls{ars}. %as a countermeasure. % AH. the score alone is not yet a countermeasure
%As a countermeasure, we develop an easy-to-use adversarial training procedure %and can confirm that it
%which significantly and successfully reduces the attack surface and improves \gls{ars}.

@@

There is a significant body of scientific work focusing on the detection of unwanted behavior in networks. In the past, a viable way of performing intrusion detection was to inspect the content of packets themselves and detect if a packet delivers potentially harmful content. More recently, with the rise of omnipresent encryption, %available features have changed. Since payload is no longer available,
the focus now lies on features that are always available to network monitoring equipment like port numbers, protocol flags or packet sizes when encrypting above the transport layer.

@@

However, for practical use, high detection accuracy is not enough. With the recent rise of interest in \gls{aml} techniques, also \gls{aml} for \glspl{ids} has been investigated. For example, \cite{bachl_walling_2019} investigates remedies for poisoning attacks on \glspl{ids} and \cite{hashemi_towards_2019} investigate adversarial robustness of common flow- and packet-based \gls{ids}.
In this research, we investigate whether adversarial samples can be found for our \gls{rnn}-based model, i.e. minimally different attack flows which are classified as benign. This is not a trivially answerable question since adversarial samples have mostly been analyzed in the context of computer vision. Our scenario significantly deviates from computer vision because \begin{enumerate*}
\item the number of features is significantly smaller and
\item only certain features can be manipulated by an adversary. % (for example the destination port cannot be manipulated)
\end{enumerate*}

@@

Several \gls{aml} methods have been proposed in the recent literature, achieving different speed-quality tradeoffs:
\cite{szegedy_intriguing_2014} shows that it is possible to create adversarial samples for an image, which look similar to the original image but are classified wrongly. \cite{goodfellow_explaining_2015} develops the \gls{fgsm}, which makes easy generation of adversarial samples possible. \cite{papernot_crafting_2016} explores the use of \gls{aml} for \glspl{rnn}, but lacks important \gls{aml} methods, which we consider in this paper.

% Max: But FGSM is high-speed... So I removed this paragraph altogether
%As finding adversarial flows is particularly difficult in our case, % for the reasons we discussed above,
%we preferred methods  producing high-quality adversarial samples while not being as fast as other methods. We implemented the following \gls{aml} methods.

>>>>

\glspl{rnn} yield attractive properties for %performing intrusion detection on network data.
constructing \glspl{ids} for network data.
With the rise of ubiquitous \gls{ml} systems, malicious actors have been catching up quickly to find new ways to exploit \gls{ml} vulnerabilities for profit. Recently developed adversarial \gls{ml} techniques focus on computer vision and their applicability to network traffic is not straightforward: Network packets expose fewer features than an image, are sequential and impose several constraints on their features. %For example, destination ports cannot be modified since that could impair communication.

We show that despite these completely different characteristics, adversarial samples can be  generated reliably for \glspl{rnn}.
To understand a classifier's potential for misclassification, we extend existing explainability techniques and propose new ones, suitable particularly % specifically crafted %AH most can also be used for non-sequential models, so lets not restrict applicability too much
for sequential data. Applying them shows that already the first packets of a communication flow are of crucial importance and are likely to be targeted by attackers. Feature importance methods show that even relatively unimportant features can be effectively abused to generate adversarial samples.
Since traditional evaluation metrics such as accuracy are not sufficient for quantifying the adversarial threat, we propose the \gls{ars} for comparing \glspl{ids}, capturing a common notion of adversarial robustness, and show that an adversarial training procedure can significantly and successfully reduce the attack surface.% and improve \gls{ars}. %as a countermeasure. % AH. the score alone is not yet a countermeasure
%As a countermeasure, we develop an easy-to-use adversarial training procedure %and can confirm that it
%which significantly and successfully reduces the attack surface and improves \gls{ars}.

@@

There is a significant body of scientific work focusing on the detection of unwanted behavior in networks. In the past, a viable way of performing intrusion detection was to inspect the content of packets themselves and detect if a packet delivers potentially harmful content. More recently, with the increasing deployment of encryption, %available features have changed. Since payload is no longer available,
the focus now lies on features that are always available to network monitoring equipment like port numbers, protocol flags or packet sizes when encrypting above the transport layer.

@@

However, for practical use, high detection accuracy is not enough. With the recent rise of interest in \gls{aml} techniques, also \gls{aml} for \glspl{ids} has been investigated. For example, \cite{bachl_walling_2019} investigates remedies for poisoning attacks on \glspl{ids} and \cite{hashemi_towards_2019} investigate adversarial robustness of common network \glspl{ids}.
In this research, we investigate whether adversarial samples can be found for our \gls{rnn}-based model, i.e. minimally different attack flows which are classified as benign. This is not a trivially answerable question since adversarial samples have mostly been analyzed in the context of computer vision. Our scenario significantly deviates from computer vision because \begin{enumerate*}
\item the number of features is significantly smaller and
\item only certain features can be manipulated if the flow should remain valid. % (for example the destination port cannot be manipulated)
\end{enumerate*}

@@

Several \gls{aml} methods have been proposed in the recent literature, achieving different speed-quality tradeoffs:
\cite{szegedy_intriguing_2014} shows that it is possible to create adversarial samples for an image, which look similar to the original image but are classified wrongly. \cite{goodfellow_explaining_2015} develops the \gls{fgsm}, which makes easy generation of adversarial samples possible. \cite{papernot_crafting_2016} explores the use of \gls{aml} for \glspl{rnn}, but lacks important \gls{aml} methods. %, which we consider in this paper.
% Max: But FGSM is high-speed... So I removed this paragraph altogether
%As finding adversarial flows is particularly difficult in our case, % for the reasons we discussed above,
%we preferred methods  producing high-quality adversarial samples while not being as fast as other methods.
We implemented the following \gls{aml} methods.
